Scaling up a GF grammar for free parsing

### 2011-12-07 doc/report/report.tex
notes from counting talbanken/saldo
report

\begin{document}
% ska gå att förstå och locka till läsning
%ska tydligt beskriva vad rapporten handlar om
%ska innehålla vettiga sökord
\title{Title of the report} %A nice titleScaling up a GF grammar and preparing it for cool stuff for Swedish}
\author{Malin Ahlberg}
\maketitle
\newpage

@@

% check references! lncs is good maybe. lnai
\newpage
\chapter{Introduction}
\section{Introduction} 
>>This work is part of bigger project, spring and next spring.
>>Connection between grammar & parser: GF, every grammar defines a parser.
>> reusablilty, freely available.
>> the big amount of data, need language technology, machine translation
Ultimate goal: want computers to 'understand' language. Not good with flat strings.
We need to translate strings of natural language to a richer structure,
a parse tree. So far there is no
freely available ruled-based parsing giving a deep analyse for Swedish.
%There are parsers which put only put part-of-speach tags to
%the text, and others that gives a deeper analyse. 
There are statistical parser, and parser which only are deal with controlled language.
The GF parser are one of those. 
%What is parsing, flat strings to trees, example picture, what is deep parsing.
So far stable and good when working with controlled languages, but recent experiments
have showed that it is also possible to parse free language.
To do this for Swedish, we need to  .. the resource grammar needed to covering
the language specific constructions. Further a large lexicon is also needed, as well
as a method for evaluation.
During this project, we have worked on and explored those things and some problems identified.
We have extended the Swedish resource grammar in GF, to give a better coverage and imported
a large lexicon. Work to compare the trees from Talbanken to GF and to translate them.

@@

Other work exist, but none is freely available. We want to create a open source
product, therefore can only use resources that allows this. Work can continue,
will not disappear, be made use of in other projects.
What we could not cover by this project, what have been covered before (in
the spring project), and why we choose those parts.

\section{Background}  
>>This work is part of bigger project, spring and next spring.
>>Connection between grammar & parser: GF, every grammar defines a parser.
To build a parser or a grammar from scratch is a heavy work. The idea for this
project is to combine already existing parts into a bigger and more expressive grammar
as well as a large scale GF lexicon.
As all GF grammars, this one can be used for parsing, and we develop it by getting
examples, ideas and test material from a treebank.
The project is hence heavily depending on three resources, which will be described
in the following sections.

@@

\subsection{Saldo}
>> syntactic information too. Mention that GF does not have this, only lexical or
>> semantic. Not 1-1.
A good parser needs a good lexicon. We have used of Saldo\cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built upon Svenskt Associations Lexicon and contains information about more than 
120 000 modern Swedish words.
The words are equipped with both semantic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of semantically connected words.\\
\newpage % only until saldo can be on same page as the pic
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

@@

\subsection{Grammatical Framework}
>> more here. First: grammar formalism with Martin-Löf.
>> snd: functional and programming language, type checking and logic: prolog (check this)
>> dependent and implicit types : agda. (read gf book and krasses book)
>> pgf\ref{}, runtime system with parsing, lin, random generation osv.
The main component of the the project is the Grammatical Framework\cite{gfbok} (GF). It is
a functional grammar formalism based on Martin-Löf type theory.
It has been developed for about 20? years and are inspired by programming languages
like Haskell\cite{haskell},
ML\cite{ml}, Prolog\cite{bratko} and
Agda\cite{agda}.
Grammars written in GF can be used together with languages 
like Java, Java script and Haskell and works for both parsing and generation.

>> The most important!!
An important feature of GF is
%Being developed with multilinguality in mind, GF makes an important 
the distinction between
\textit{abstract} and \textit{concrete} syntaxes. The abstract is meant to be
shared by many languages, it models the semantics of the domain without concern for
language specific features such as agreement or word order.

\begin{figure}[h]
\begin{verbatim}
 abstract TestGrammar = {
  cat N ; V ; S ;

@@

>>not very restricted, not few ambiguities.
>>the scope is controlled language, but possible ambiguities (discuss this in future
>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
>>syntactical (man with the telescope), 
>>lexical (youSg/Pl, växa).
A GF grammar usually covers a very restricted set of language where few ambiguities
occur and much care can be taken to preserving the semantics of the
different languages.
By using a controlled language like this,  good and reliable translation can be
ensured. Some examples of projects using GF are WebAlt\cite{webalt},
a project which aims to develop language independent material for mathematics problems,
and the formal software verification tool KeY\cite{key}, in wich GF is used for translating specification
to natural language.
>> key, talk, hats, bom (see references in Ramonas thesis).
%, the Online Syllogism Solver\cite{malin!}. 
The framework is also used in the
European project MOLTO\cite{molto} for online translation between up to 15 language.\\

\subsubsection{The resource library}
>>Best part? Biggest library?
>>Most general resource, first large grammar, covers fundamental syntactics.
The resource library\cite{gf-resource} is a very useful part of the framework. It is provided in the 
GF package. %, they give a good start for all who want to write an application grammar.
Instead of letting the users of the framework reimplement the morphological and
the fundamental syntactic structures of a language for each application, all
this can be found as libraries. So far there are 20 languages in the library, and
they all share the same abstract syntax.
There are also a small test lexicon included, containing some hundred of common words.

@@

\vspace{5mm}
>> references here! to everything. Remove uncertain, add ontology, aarne paper about
>> donkey, find aarnes paper in ramona's thesis (french), hats formal methods,
>> bom.
As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of använad det på ett
lätt sätt are added and the possibilities to use GF in different environments
increased. There are  research about how to make better use of the dependent
types, for solving syllogism or generating natural language via Montague
semantics\cite{montague}.
Further, experiments on using GF for free parsing has been conducted
\cite[\textsection 4.1]{gfMech} 
This is also a future goal for our project and one of the delmål has been to
model an important part of Swedish in GF. The starting point was the resource
grammar and the result should still have a strong connection to this.

%Functor
%Main ideas, abstract and concrete, translation, parse- and abstract trees.
%Example of abstract.
%We say that this is a function that takes as argument and returns..
%Categories, valency built in (problems with ställa - på, under, vid). Parameters, fields.
%Example of a piece with Swedish, the reader gets familiar with how it looks.
%Also show how information 'disappears', become strings.
%What kind of linguistic analyse, what is the purpose to cover. Mention other uses,
%ideas of anaphores, montague? dependent types, ontologies, cool pictures by Krasse.
%
%The resources, why, how many, how to use, projects. functors.
%
%Still under big development, evolving, growing.
%For this project, only (mostly) important to model Swedish without abandoning the
%resources too much. Use of the big grammar for translation, why it might be hard with
%semantics and lexicon.
%Research about how to parse with gf by Krasimir, experiments with English.

@@

\subsection{Talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being an freely available, manually annotated, large scale treebank,
>> cite mamba? annotation: well intergrated in sw. tradition, descriptive, used
>> for large scale projects.
Talbanken\cite{talbanken} was perfect for our purpose.
It was created in the 1970s at Lund University. In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school students, and the last one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser\cite{malt}
and was thus redistributed in a more modern version\cite{talbanken05} This work
was done by Nivre, Hall and Nilsson in 2005. They released it in Malt and Tiger XML-formats,
where the trees had been made deeper and more detailed. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Related work}
>> Much more here, important!! comparisons.
Computational ling is active area for Swedish, there are many other Swedish parser,
but grammars?
Many of the parsers not ruled based, the most known one Malt\cite{malt} ... .
However has been two, about CLE\cite{cle} and what happened to it. Logic and translation
to Swedish, S-CLE (Gambäck,1997) full analyse but coverage is too limited (nivres papper), The
spoken lang. translator\cite{cle2}. Wiren \cite{wiren}. bit more information than
just pos: cassSwe \cite{casswe}
Other Sw. parsers and grammars, evaluation of them? 
Other grammar frameworks.
SweCG (birn) annotation, parser Swedish FDG (voultanien,2001) which is not freely available.
finding subject, objects, suubject complements (SC) and
to link them to their proper regents (main verbs) : over 90\% both precision and recall.
uses a dependency grammar, but "The functional description of adverb phrases
and prepositional phrases (e.g. agent, source, goal, benefactive, time) remains
to be described in a future version." 
HPSG\cite{hpsg} and lingo matrix\cite{lingomatrix}, English, Japanese, German,
Greece, Spanish, Norweigan (NorSource, Hellan 2003),( not famous: SweCore
(Ahrenberg) based on NorSource)
also Swedish: BiTSE\cite{stymne}, Stymne 2006, Swedish English translation,
focusing on verb phrames, built on HPSG.
S\o rgaard and Haugereid, Scandinavian Grammar Matrix\cite{scandmatrix}
Tag, Tree adjoining grammar, XTAG: English and Korean, Chinese? English: ongoing to get widecoverage grammar
under GPL. Translation to Korean.
% related and references work: CLE, Krasimirs book, CassSwe, GF book,SAG, gunlög,
% the multilingual treebank, Homse&Hinchcliff, Aarne i LILT

@@

\section{Importing Saldo}
\label{sec:prog.saldo}
Saldo is compatible with GF, which makes it easy to transform the morphological
lexicon into our format. The lexicon has been imported earlier (Krasimir,2009?)
but the code needed updates to not miss many important words.
Experiments have been made to use a interactive tool for lexical acquisition,
but the dictionary was still much to small. The acquisition tool may still
be used for complementing the lexicon.

@@

The resulting dictionary contains more than 100 000 entries, compared to the total size
of Saldo this is approximately 80\%. When testing the coverage of Talbanken,
we found that 3058 words were still missing. In addition, there were 1817 words
which were given different labels in GF than in Talbanken. Many of those are
acceptable, such as the ones in table \ref{tab:saldodiff} and reflects the difference 
made in the analyses, but others are examples of words that are still missing from
the lexicon.
\begin{figure}[h]
\begin{tabular}{lll}
Word & Talbanken tag & GF category \\
\hline
måste & MVPS          & VV \\
allting & POTP & N \\
få & POZP & Det \\
\end{tabular}
\caption{}
\label{tab:saldodiff}
\end{figure}
Among the words that were completely missing, more than 2000 was found to be nouns,
many of which were compounds.

@@

\subsection{Added}
\label{sec:Added}
Summary of what have been added and motivation.

@@

\section{Mapping}
>> no subjects or object, analyse is more from the lexicon,
>> compare to others, ask elisabet.
\label{sec:Mapping}
Talbanken contains much valuable information about phrase structures and word 
usage. One part of this project has focused on extracting this information
and to translate it to GF notation. We have constructed a mapping, which 
automatically transforms trees in Talbanken format to GF abstract syntax trees.
Figure \ref{fig:translationtrees} shows an example of a visualized tree
from Talbanken05 and the result of translating it to GF.
Both the POS tags as well as the syntactic information are needed for the translation.

>> add parse tree, and describe connection between parse and abstract
>> show how we keep analyse
\begin{figure}[h]
\centering
\subfloat[Talbanken tree]{\label{pic:tbtree}\includegraphics[width=60mm]{Talbankentree.pdf}}
\subfloat[GF abstract tree]{\label{pic:gftree}\includegraphics[width=60mm]{gfTree.pdf}}
\caption{Trees for the sentence ``Katten på bilen blir större".}
\label{fig:translationtrees}
\end{figure}

@@

A translation of the English Penn Treebank to the GF syntax has previously been done \cite{gfpenn}.
By starting this from program, we now have
a translation that works for the quite differently annotated Talbanken. The differences between
Swedish and English syntax also meant that many rules had to be changed.\\

The mapping works for Talbanken05 in Tiger XML-format. There are three kinds of
tags: categories, edge labels and pos tags. 
The categories give syntactic information such as \verb|S|,\verb|NP| and \verb|VP|.
The edge labels give similar information with tags as \verb|SS| for subject, 
\verb|OO| for object. There are more than 300 different pos tags. The high
number is due to the detailedness, for example there are almost 50 tags for
nouns, excluding proper names. The tags show definiteness, case, whether the
word is a compound. Some words also have their own tags.\\

\begin{tabular}{lll}
\textbf{Tag} & \textbf{Example} & \textbf{Explanation} \\
\hline
\textbf{SV} & ska & The verb \emph{shall}\\
\textbf{WVIV} & vilja & The verb \emph{want} in infinitive\\
\textbf{KVSP} & kommit att & The verb \emph{will} in supine\\
\textbf{GVPT  PA} & gjordes & The verb \emph{do, make}, present, passive \\
\textbf{NNDDHSGG} & familjemedlemmarnas & Noun, definite, compound (person), genitive \\
\textbf{PU} & *, 1), a) & List item  ...\\
\end{tabular}\\

@@

The translation works section-wise, ie, if it finds a tree of category \verb|S|,
it will start to look for a subject \verb|SS| and a finite verb \verb|FV|. If it
can find those, they can be combine with the GF function \verb|PredVP|.
At the lowest level in the tree we look up the word in the GF lexicon.
It figure \ref{pic:tbtree}, the verb is followed by an adjective as a present
participle, \verb|SP|. GF admits \emph{bli} to have type \verb|VA|, a verb
with an adjective as complement, and the translator can apply the gf function
\verb|ComplVA| to combine the verb with its object. \\
It is not often the case that we can first find the subject, then the verb and 
finally the objects. Often the order is Verb-Subject-Object or Object-Subject-Verb. \\
During the mapping process, the translator also needs to find out whether the sentence is
negated and what tense is used. If it finds two trees of type, say, \verb|VP| but does not
have any rules for combining them, the \textit{meta function} \emph{'?'} will be used. 
Figure xx shows an example of this.\\

@@

\section{Conclusion}
>> why is this significant, other grammars can use this Swedish grammar.
>> well known (comp. ling) trade-off between depth & coverage. for now depth, robustness in in future
>> work leads to coverage.
%There is obviously a lot do still, will never get finish with NPL.
%Grammar may need even bigger changes, to allow enough but not too much.
%Different from writing a grammar for generation.
%Results from mapping shows that the translation is doable, but becomes harder
%since talbanken don't have formal rules.
%The saldo shows how easy to make use of utomstående resources, promising if
%we want to extract other information. 
%Hard, but not finished!

\bibliographystyle{ieeetr}
\bibliography{masterbib}
\end{document}

>>>>

\begin{document}
% ska gå att förstå och locka till läsning
%ska tydligt beskriva vad rapporten handlar om
%ska innehålla vettiga sökord
\title{Scaling up a GF grammar for free parsing} %A nice titleScaling up a GF grammar and preparing it for cool stuff for Swedish}
\author{Malin Ahlberg}
\maketitle
\newpage

@@

% check references! lncs is good maybe. lnai
\newpage
\chapter{Introduction}
\section{Introduction} 
>>This work is part of bigger project, spring and next spring.
>>Connection between grammar and parser: GF, every grammar defines a parser.
>> reusablilty, freely available.
>> the big amount of data, need language technology, machine translation
>> parsing: assingn of morph. and syntactic structures.
>>Connection between grammar and parser: GF, every grammar defines a parser.
Ultimate goal: want computers to 'understand' language. Not good with flat strings.
We need to translate strings of natural language to a richer structure,
a parse tree. So far there is no
freely available ruled-based parsing giving a deep analyse for Swedish.
%There are parsers which put only put part-of-speach tags to
%the text, and others that gives a deeper analyse. 
There are statistical parser, and parser which only are deal with controlled language.
The GF parser are one of those. 
%What is parsing, flat strings to trees, example picture, what is deep parsing.
So far stable and good when working with controlled languages, but recent experiments
have showed that it is also possible to parse free language.
To do this for Swedish, we need to  .. the resource grammar needed to covering
the language specific constructions. Further a large lexicon is also needed, as well
as a method for evaluation.
During this project, we have worked on and explored those things and some problems identified.
We have extended the Swedish resource grammar in GF, to give a better coverage and imported
a large lexicon. Work to compare the trees from Talbanken to GF and to translate them.

@@

Other work exist, but none is freely available. We want to create a open source
product, therefore can only use resources that allows this. Work can continue,
will not disappear, be made use of in other projects.
%What we could not cover by this project, what have been covered before (in
%the spring project), and why we choose those parts.
To build a parser or a grammar from scratch is a heavy work and we
therefore combine already existing parts.

\section{Background}  
The work described in this is thesis is part of a bigger project which aims at
using GF for parsing free Swedish. 
In previous work \ref{??} the start on extending the Swedish GF grammar was
made and a tool for lexical aquistion were developed.
We now construct a bigger
and more expressive grammar as well as a large scale GF lexicon.
As all GF grammars, this one can be used for parsing, and we develop it by getting
examples, ideas and test material from a treebank.
The project is hence heavily depending on three resources, which will be described
in this section.

@@

\subsection{Saldo}
A good parser needs a good lexicon. We have used of Saldo\cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built upon Svenskt Associations Lexicon and contains information about more than 
120 000 modern Swedish words.
The words are equipped with semantic, syntactic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of semantically connected words.\\
\newpage % only until saldo can be on same page as the pic
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

@@

\subsection{Grammatical Framework}
The main component of the the project is the Grammatical Framework\cite{gfbok} (GF). It is
a grammar formalism based on Martin-Löf type theory\cite{martinlof}. GF is
designed for multilingual applications and is equivalent to Parallel Multiple
Context-Free Grammars (Seki \& al. 1991), it is hence stronger than mildly
contex-free grammars.

GF is a functional programming languages, inspired by programming languages
like ML\cite{ml}, Haskell\cite{haskell}, Prolog\cite{bratko} and
Agda\cite{agda}. 
It has been developed for about 20 years 
>> snd: functional and programming language, type checking and logic: prolog (check this)
>> dependent and implicit types : agda. (read gf book and krasses book)

A grammar written in GF can be used for both parsing and generation.
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system. The interpreter, the GF shell, lets
you try and test your grammar by using commands for parsing, vizualization of parse trees,
random generation, word alignment, morphological quizzes etc.
The compiler lets you compile your grammar into portable grammar format (PGF)\ref{pgf}.
There are libraries for using PGF files with Java, Java script and Haskell. The grammars
may also be used in Android applications.
and are inspired by.\\

The key feature in GF is the distinction between
\textit{abstract} and \textit{concrete} syntaxes. The abstract is meant to be
shared by many languages and can be implemented by a numeber of concrete grammars,
each representing one language. The abstract syntax models the semantics of the
domain without concern for language specific features such as agreement or word
order.

%Being developed with multilinguality in mind, GF makes an important 
\begin{figure}[h]
\begin{verbatim}
 abstract TestGrammar = {
  cat N ; V ; S ;

@@

>>not very restricted, not few ambiguities.
>>the scope is controlled language, but possible ambiguities (discuss this in future
>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
>>syntactical (man with the telescope), 
>>lexical (youSg/Pl, växa).
A GF grammar usually covers a very restricted set of language where few ambiguities
occur and much care can be taken to preserving the semantics of the
different languages.
By using a controlled language like this,  good and reliable translation can be
ensured. Some examples of projects using GF are WebAlt\cite{webalt},
a project which aims to develop language independent material for mathematics problems,
and the formal software verification tool KeY\cite{key}, in wich GF is used for
translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%, the Online Syllogism Solver\cite{malin!}. 
The framework is also used in the
European project MOLTO\cite{molto} for online translation between up to 15 language.\\

\subsubsection{The resource library}
%>>Best part? Biggest library?
%>>Most general resource, first large grammar, covers fundamental syntactics.
The GF package provides a powerful resource library\cite{gf-resource}, which covers the
fundamental morphology and syntactics for more than 20 languages. %is a very useful part of the framework. It is provided in the 
%, they give a good start for all who want to write an application grammar.
%of a language for each application, all
%this can be found as libraries. So far there are 20 languages in the library, and
The languages all share the same abstract and they constitute  
a very useful resource when implementing application grammars.
There is also a small test lexicon included, containing some hundred of common
words.

@@

A GF parse tree has two visualized representations, the \textit{parse tree} and
the \textit{abstract tree}, as shown in figure \ref{fig:gfPandAtrees}.
>> have other treees! One in english.
\begin{figure}[h]
\centering
\subfloat[GF parse tree]{\label{pic:gfPtree}\includegraphics[width=50mm]{gfPTree.pdf}}
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=60mm]{gfTree.pdf}}
\caption{Parse tree and abstract tree for the sentence ``Katten på bilen blir större".}
\label{fig:gfPandAtrees}
\end{figure}
The abstract trees shows all functions GF needs to use
to parse the sentence and is consequently the same for all concrete syntaxes.
The parse tree (\ref{pic:gftree}) shows the words and which types the functions assign
to them.
The abstract tree shows us that the function
\verb_DetCN_ is applied the \verb-UseN cat_N- and the information
definitive (\verb-DefArt-) and singular (\verb-NumSg-). In the Swedish parse tree
we see that \emph{``katten"}, which is in definite form singular, is noun \verb-N-
which is turned into a \verb-CN- and finally a \verb-NP-. In the English parse
tree, the form of \emph{``cat"} expresses singular while the definite article
\emph{``the"} is explicit and therefore shown in the parse tree.

\vspace{5mm}
As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of using it in more 
effecient and easier ways are added
and the possibilities to use GF in different environments
increased. There are  research about how to make better use of the dependent
types, for using ontologies\cite{ramona}(>>check!!) or generating natural
language via Montague semantics\cite{montague}.
Further, experiments on using GF for free parsing has been conducted
\cite[\textsection 4.1]{gfMech}.
This is also a future goal for our project and one of the subgoal has been to
model an important part of Swedish in GF. The starting point was the resource
grammar and the result should still have a strong connection to this.

%Functor
%Main ideas, abstract and concrete, translation, parse- and abstract trees.
%We say that this is a function that takes as argument and returns..
%Categories, valency built in (problems with ställa - på, under, vid). Parameters, fields.
%Example of a piece with Swedish, the reader gets familiar with how it looks.
%Also show how information 'disappears', become strings.
%What kind of linguistic analyse, what is the purpose to cover. Mention other uses,
%ideas of anaphores, montague? dependent types, ontologies, cool pictures by Krasse.
%
%The resources, why, how many, how to use, projects. functors.
%
%Still under big development, evolving, growing.
%For this project, only (mostly) important to model Swedish without abandoning the
%resources too much. Use of the big grammar for translation, why it might be hard with
%semantics and lexicon.
%Research about how to parse with gf by Krasimir, experiments with English.

@@

\subsection{Talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being an freely available, manually annotated, large scale treebank,
Talbanken\cite{talbanken} was perfect for our purpose.
It was created in the 1970s at Lund University. In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school students, and the last one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser\cite{malt}
and was thus redistributed in a more modern version\cite{talbanken05} This work
was done by Nivre, Hall and Nilsson in 2005. They released it in Malt and Tiger
XML-formats\footnote{http://www.ims.uni-stuttgart.de/projekte/TIGER/TIGERSearch/doc/html/TigerXML.html},
where the trees had been made deeper and more detailed. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Related work}
>> Much more here, important!! comparisons.
Computational ling is active area for Swedish, there are many other Swedish parser,
but grammars?
Many of the parsers not ruled based, the most known one Malt\cite{malt} ... .
However has been two, about CLE\cite{cle} and what happened to it. Logic and translation
to Swedish, S-CLE (Gambäck,1997) full analyse and semantics, translation,
but coverage is too limited (nivres papper), The
spoken lang. translator\cite{cle2}. Wiren \cite{wiren}?? chartparser, but for swedish. bit more information than
just pos: cassSwe \cite{casswe}
Other Sw. parsers and grammars, evaluation of them? 
Other grammar frameworks.
SweCG (birn) annotation, parser Swedish FDG (voultanien,2001) which is not freely available.
finding subject, objects, suubject complements (SC) and
to link them to their proper regents (main verbs) : over 90\% both precision and recall.
uses a dependency grammar, but "The functional description of adverb phrases
and prepositional phrases (e.g. agent, source, goal, benefactive, time) remains
to be described in a future version." 
HPSG\cite{hpsg} and lingo matrix\cite{lingomatrix}, English, Japanese, German,
Greece, Spanish, Norweigan (NorSource, Hellan 2003),( not famous: SweCore
(Ahrenberg) based on NorSource)
also Swedish: BiTSE\cite{stymne}, Stymne 2006, Swedish English translation,
and one for the common parts of scandinavian (Haugereid etc) in Nordisk sprog.
focusing on verb phrames, built on HPSG.
S\o rgaard and Haugereid, Scandinavian Grammar Matrix\cite{scandmatrix}
Tag, Tree adjoining grammar, XTAG: English and Korean, Chinese? English: ongoing to get widecoverage grammar
under GPL. Translation to Korean.
% related and references work: CLE, Krasimirs book, CassSwe, GF book,SAG, gunlög,
% the multilingual treebank, Homse&Hinchcliff, Aarne i LILT

@@

\section{Importing Saldo}
\label{sec:prog.saldo}
>> syntactic information too. Mention that GF does not have this, only lexical or
>> semantic. Not 1-1.
Saldo is compatible with GF, which makes it easy to transform the morphological
lexicon into our format. The lexicon has been imported earlier (Krasimir,2009?)
but the code needed updates to not miss many important words.
Experiments have been made to use a interactive tool for lexical acquisition,
but the dictionary was still much to small. The acquisition tool may still
be used for complementing the lexicon.

@@

The resulting dictionary contains more than 100 000 entries, compared to the total size
of Saldo this is approximately 80\%. When testing the coverage of Talbanken,
we found that around 3000 words were still missing (excluding the one tagged as
numerals and proper names, but some names are still there). In addition, there were 1817 words
which were given different labels in GF than in Talbanken. Many of those are
acceptable, such as the ones in table \ref{tab:saldodiff} and reflects the difference 
made in the analyses, but others are examples of words that are still missing from
the lexicon.
\begin{figure}[h]
\begin{tabular}{lll}
Word & Talbanken tag & GF category \\
\hline
måste & MVPS          & VV \\
allting & POTP & N \\
få & POZP & Det \\
\end{tabular}
\caption{}
\label{tab:saldodiff}
\end{figure}
Among the words that were completely missing, more than 2000 was found to be nouns,
many of which were compounds.

@@

\subsection{Development of the grammar}
\label{sec:Added}
Summary of what have been added and motivation.

@@

\section{Mapping}
>> no subjects or object, analyse is more from the lexicon,
>> compare to others, ask elisabet.
>> diffs: valencies, pp (gf ingen skillnad direkt/indirekt i träd, pga multiling)
>> mamba very used and tested for swedish, standard.
\label{sec:Mapping}
Talbanken contains much valuable information about phrase structures and word 
usage. It is analysed with the MAMBA annotation scheme (Telemann, 1974) which is 
%>> cite mamba? annotation: well intergrated in sw. tradition, descriptive, used
%>> for large scale projects.
%annotation scheme
commonly used, and hence very well-tested, for describing Swedish.
The deepened analyse in Talbanken05 includes the lexical MAMBA layer. 
One part of this project has focused translating trees from Talbanken to GF.
We hence get a comparision between the two annotaitons and at the same time
we get ways to extract and translate information to GF notation. We have
constructed a mapping, which 
automatically transforms trees in Talbanken format to GF abstract syntax trees.
Figure \ref{fig:translationtrees} shows an example of a visualized Talbanken05 tree
of the sentence \emph{``The cat on the car gets bigger"} and the result of translating it to GF.
During the translation all the information given in the abstract tree
(\ref{pic:gftree}) needs to be extracted. We need to know that \emph{``katten"} is a noun
used in definite form (\verb-DetArt-), singular (\verb-NumSg-). Even though the
notation is changed by the transformation, the original analyse from Talbanken,
that ``på bilen" forms a subtree which is attatched to ``katten" for example,
is preserved.
%produces a \verb-NP- from ``katten" this means that
%DetCN : Quant                  -> CN    -> NP
%DetCN  (DetQuant DefArt NumSg)    cat_N = katten
Both the POS tags as well as the syntactic information are needed.

\begin{figure}[h]
\hspace{-10mm}
\subfloat[Talbanken tree]{\label{pic:tbtree}\includegraphics[width=60mm]{Talbankentree.pdf}}
\subfloat[GF abstract tree]{\label{pic:gftree}\includegraphics[width=60mm]{gfTree.pdf}}
\subfloat[GF parse tree]{\label{pic:gfptree}\includegraphics[width=50mm]{gfPTree.pdf}}
\caption{Trees for the sentence ``Katten på bilen blir större".}
\label{fig:translationtrees}
\end{figure}

@@

A translation of the English Penn Treebank to the GF syntax has previously been done \cite{gfpenn}.
By starting this from program, we now have
a translation that works for the differently annotated Talbanken. The differences between
Swedish and English syntax also meant that many rules had to be changed.\\

The mapping works for Talbanken05 in Tiger XML-format.
There are three kinds of tags: categories, edge labels and POS tags. 
The categories give syntactic information such as \verb|S|, \verb|NP| and \verb|VP|.
The edge labels give similar information with tags as \verb|SS| for subject, 
\verb|OO| for object. There are more than 300 different pos tags. The high
number is due to the detailedness, for example there are almost 50 tags for
nouns, excluding proper names. The tags show definiteness, case, whether the
word is a compound. Some words also have their own tags.\\

\begin{tabular}{lll}
\textbf{Tag} & \textbf{Example} & \textbf{Explanation} \\
\hline
\textbf{SV} & ska & The verb \emph{shall}\\
\textbf{WVIV} & vilja & The verb \emph{want} in infinitive\\
\textbf{KVSP} & kommit att & The verb \emph{will} in supine\\
\textbf{GVPT  PA} & gjordes & The verb \emph{do, make}, present, passive \\
\textbf{NNDDHSGG} & familjemedlemmarnas & Noun, definite, compound (person), genitive \\
\textbf{PU} & * $\;$ 1) $\;$ a) & List item\\
\end{tabular}\\

@@

The translation works section-wise, ie. if it finds a tree of category \verb|S|,
it will start to look for a subject \verb|SS| and a finite verb \verb|FV|. If it
can find those, they can be combine with the GF function \verb|PredVP|.
At the lowest level in the tree we look up the word in the GF lexicon.
It figure \ref{pic:tbtree}, the verb is followed by an adjective as a present
participle, \verb|SP|. GF admits \emph{``bli"} to have type \verb|VA|, a verb
with an adjective as complement, and the translator can apply the GF function
\verb|ComplVA| to combine the verb with its object. \\
It is not always the case that we first find the subject, then the verb and 
finally the objects. Often the order is Verb-Subject-Object or Object-Subject-Verb. \\
During the mapping process, the translator also needs to find out whether the sentence is
negated and what tense is used. If it finds two trees of type, say, \verb|VP| but does not
have any rules for combining them, the \textit{meta function} \verb-?- will be used. 
Figure xx shows an example of this.\\

@@

\section{Conclusion}
>> why is this significant, other grammars can use this Swedish grammar.
>> well known (comp. ling) trade-off between depth and coverage. for now depth, robustness in in future
>> work leads to coverage.
%There is obviously a lot do still, will never get finish with NPL.
%Grammar may need even bigger changes, to allow enough but not too much.
%Different from writing a grammar for generation.
%Results from mapping shows that the translation is doable, but becomes harder
%since talbanken don't have formal rules.
%The saldo shows how easy to make use of utomstående resources, promising if
%we want to extract other information. 
%Hard, but not finished!

\bibliographystyle{lncs}
\bibliography{masterbib}
\end{document}
